Delayed effects of climate on vital rates leads to demographic divergence in Amazonian forest fragments

### 2021-05-13 doc/paper.Rmd
added one citation

<!--# what do you think of this wording? Trying to walk people through the big result by breaking into two sentences telling them what they are about to see --> For 1-ha fragments, recent SPEI values (i.e. within the past 0--12 months) had the strongest effect on survival <!--# Is another way to say this "For 1-ha fragments, SPEI in the preceding 12 months had the strongest effect on survival, with..."  -->., with the highest survival near SPEI of 0 and mortality increasing as conditions became either drier or wetter (i.e., as SPEI values became increasingly negative or positive, respectively; Figure \@ref(fig:surv-spei)a).
<!--# what do you think? trying to put the results in more ecological terms) -->In contrast, the effect of recent SPEI in continuous forest was both weaker and unidirectional -- survival probability was reduced only when SPEI was \< -1 (i.e., mild-severe drought), and even then only slightly (Figure \@ref(fig:surv-spei)b).
Instead the most pronounced effects on the survival of plants in continuous forest were of SPEI at a lag time of 15--20 months (i.e. two dry seasons prior to a census).
Drought 15--20 months prior to a census (i.e., SPEI \< -1) was associated with reduced survival, while high precipitation (i.e., SPEI \> 1) was associated with higher survival (put another way, was a nearly linear relationship between SPEI and survival probability).
Finally, plants in both habitat types showed an increase in survival probability with very high SPEI values (i.e., extremely high precipitation) at a lag time of 36 months.
It should be noted, however, that only the first year of census data (1999) met these conditions.
<!--# can you walk me through these next two sentences? I'm not sure if this applies to the 36 mop lag or the whole time-series -->Subtracting the estimated contribution of SPEI history to survival in fragments from that in continuous forest shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and fragments (Figure \@ref(fig:surv-spei)c).
However, under extreme conditions, survival probability is higher in continuous forest by up to 0.02.

@@

Climate models for the Amazon predict a future of extremes---increases in the frequency and geographic extent of droughts, but also increases in the frequency and area affected by periods of unusual wetness [@duffy2015]. Our results support the hypothesis that populations in fragments could be more susceptible to the effects of changing climate than those in continuous forest [@lauranceApparentEnvironmentalSynergism2014]. However, they also indicate that the demographic responses to climate change of populations in fragmented landscapes may be far more complex than previously appreciated. Multi-factorial, multi-season experiments are needed to determine if habitat-specific differences in environmental conditions and local ecological neighborhood interact to delay the demographic responses of plants to climatic variability and the mechanisms underlying these responses [*sensu* @brunaCompensatoryResponsesUnderstory2005; @markewitz2010; @westerbandInteractionsPlantSize2015; @westerbandEarlyLifeConditions2017]

Also needed are statistical tools that can test for synergistic effects of fragmentation and climate in vital rates, as those currently available do not allow for including interaction terms.
This also limits the ability to include size by climate interactions in a DLNM; although plant responses to both fragmentation and climatic extremes can be size-specific [@brunaDemographicEffectsHabitat2005; @schwartzFragmentationForestStructure2019].
The ability to identify size-specific lagged responses may be especially complicated given size and growth are rarely measured at the same time scale as SPEI and other putative climatic drivers.

>>>>

<!--# what do you think of this wording? Trying to walk people through the big result by breaking into two sentences telling them what they are about to see -->

For 1-ha fragments, recent SPEI values (i.e. within the past 0--12 months) had the strongest effect on survival <!--# Is another way to say this "For 1-ha fragments, SPEI in the preceding 12 months had the strongest effect on survival, with..."  -->., with the highest survival near SPEI of 0 and mortality increasing as conditions became either drier or wetter (i.e., as SPEI values became increasingly negative or positive, respectively; Figure \@ref(fig:surv-spei)a).
<!--# what do you think? trying to put the results in more ecological terms) -->In contrast, the effect of recent SPEI in continuous forest was both weaker and unidirectional -- survival probability was reduced only when SPEI was \< -1 (i.e., mild-severe drought), and even then only slightly (Figure \@ref(fig:surv-spei)b).
Instead the most pronounced effects on the survival of plants in continuous forest were of SPEI at a lag time of 15--20 months (i.e. two dry seasons prior to a census).
Drought 15--20 months prior to a census (i.e., SPEI \< -1) was associated with reduced survival, while high precipitation (i.e., SPEI \> 1) was associated with higher survival (put another way, was a nearly linear relationship between SPEI and survival probability).
Finally, plants in both habitat types showed an increase in survival probability with very high SPEI values (i.e., extremely high precipitation) at a lag time of 36 months.
It should be noted, however, that only the first year of census data (1999) met these conditions.
<!--# can you walk me through these next two sentences? I'm not sure if this applies to the 36 mop lag or the whole time-series -->Subtracting the estimated contribution of SPEI history to survival in fragments from that in continuous forest shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and fragments (Figure \@ref(fig:surv-spei)c).
However, under extreme conditions, survival probability is higher in continuous forest by up to 0.02.

@@

Climate models for the Amazon predict a future of extremes---increases in the frequency and geographic extent of droughts, but also increases in the frequency and area affected by periods of unusual wetness [@duffy2015]. Our results support the hypothesis that populations in fragments could be more susceptible to the effects of changing climate than those in continuous forest [@lauranceApparentEnvironmentalSynergism2014]. However, they also indicate that the demographic responses to climate change of populations in fragmented landscapes may be far more complex than previously appreciated. Multi-factorial, multi-season experiments [*sensu* @brunaCompensatoryResponsesUnderstory2005; @markewitz2010; @westerbandEarlyLifeConditions2017; @aguirre2021], ideally manipulating multiple combinations of climatic variables [@mundim2016], are needed to determine how and why habitat-specific differences in environmental conditions interact to delay the demographic responses of plants to climatic variability. Also needed are statistical tools that can test for synergistic effects of fragmentation and climate in vital rates, as those currently available do not allow for including interaction terms. This also limits the ability to include size by climate interactions in a DLNM; although plant responses to both fragmentation and climatic extremes can be size-specific [@brunaDemographicEffectsHabitat2005; @schwartzFragmentationForestStructure2019].
The ability to identify size-specific lagged responses may be especially complicated given size and growth are rarely measured at the same time scale as SPEI and other putative climatic drivers.
